---
title: "Recluster fibroblast lineage"
author: "Belinda Phipson & Neda R. Mehdiabadi"
date: "10/28/2019-05/11/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE}
knitr::opts_chunk$set(error = TRUE)
```

# Load libraries and functions

```{r}
library(edgeR)
library(RColorBrewer)
library(org.Hs.eg.db)
library(limma)
library(Seurat)
library(cowplot)
library(DelayedArray)
library(scran)
library(NMF)
library(workflowr)
library(ggplot2)
library(clustree)
library(dplyr)
```

```{r}
source("code/normCounts.R")
source("code/findModes.R")
source("code/ggplotColors.R")
```

```{r readTargets}
targets <- read.delim("data/targets.txt",header=TRUE, stringsAsFactors = FALSE)
targets$FileName2 <- paste(targets$FileName,"/",sep="")
targets$Group_ID2 <- gsub("LV_","",targets$Group_ID)
group <- c("fetal_1","fetal_2","fetal_3",
           "non-diseased_1","non-diseased_2","non-diseased_3",
           "diseased_1","diseased_2",
           "diseased_3","diseased_4")
m <- match(group, targets$Group_ID2)
targets <- targets[m,]
```

```{rm readData}
fetal.integrated <- readRDS(file="/group/card2/Neda/MCRI_LAB/single_cell_nuclei_rnaseq/Porello-heart-snRNAseq/output/RDataObjects/fetal-int.Rds")
load(file="/group/card2/Neda/MCRI_LAB/single_cell_nuclei_rnaseq/Porello-heart-snRNAseq/output/RDataObjects/fetalObjs.Rdata")
##note: nd.integrated is also an integrated form of young heart samples. young.integrated has already been published in Sim et al., 2021
nd.integrated <- readRDS(file="/group/card2/Neda/MCRI_LAB/single_cell_nuclei_rnaseq/Porello-heart-snRNAseq/output/RDataObjects/nd-int.Rds")
load(file="/group/card2/Neda/MCRI_LAB/single_cell_nuclei_rnaseq/Porello-heart-snRNAseq/output/RDataObjects/ndObjs.Rdata")

dcm.integrated <- readRDS(file="/group/card2/Neda/MCRI_LAB/single_cell_nuclei_rnaseq/Porello-heart-snRNAseq/output/RDataObjects/dcm-int.Rds")
load(file="/group/card2/Neda/MCRI_LAB/single_cell_nuclei_rnaseq/Porello-heart-snRNAseq/output/RDataObjects/dcmObjs.Rdata")
```

# Set default clustering resolution

```{rm}
# Default 0.3
Idents(fetal.integrated) <- fetal.integrated$integrated_snn_res.0.3
DimPlot(fetal.integrated, reduction = "tsne",label=TRUE,label.size = 6)+NoLegend()

# Default 0.3
Idents(nd.integrated) <- nd.integrated$integrated_snn_res.0.3
DimPlot(nd.integrated, reduction = "tsne",label=TRUE,label.size = 6)+NoLegend()

# Default 0.3
Idents(dcm.integrated) <- dcm.integrated$integrated_snn_res.0.3
DimPlot(dcm.integrated, reduction = "tsne",label=TRUE,label.size = 6)+NoLegend()

```

# Merge all data together

```{r}
# This data has already been generated and saved as the heart object.
#heart <- merge(fetal.integrated, y = c(nd.integrated, dcm.integrated), project = "heart")
heart <- readRDS("/group/card2/Neda/MCRI_LAB/must-do-projects/EnzoPorrelloLab/dilated-cardiomyopathy/data/heart-int-FND-filtered.Rds")
table(heart$orig.ident)
```

# Get fibroblast cells only

```{r}
Idents(heart) <- heart$Broad_celltype
fibro <- subset(heart,subset = Broad_celltype == "Fib")
```

# Check for poor quality cells

Check for cells with very low number of uniquely detected genes.

```{r}
DefaultAssay(fibro) <- "RNA"
par(mar=c(4,4,2,1))
plot(density(fibro$nFeature_RNA),main="Number of genes detected")
abline(v=500,col=4, lty=3)
legend("topright",lty=2,col=4,legend="#genes = 500")
plot(density(fibro$nCount_RNA),main="Library size")
abline(v=2500,col=4, lty=3)
legend("topright",lty=2,col=4,legend="library size = 2500")

fibro <- subset(fibro, subset = nFeature_RNA > 500 & nCount_RNA > 2500)
```


# Run new integration with SCtransform normalisation


```{rm}
# For the sake of time, I ran the following code once and saved the object as fibro-int-FND.Rds

fibro.list <- SplitObject(fibro, split.by = "biorep")
min <-  min(sapply(fibro.list, ncol))
for (i in 1:length(fibro.list)) {
    fibro.list[[i]] <- SCTransform(fibro.list[[i]], verbose = FALSE)
}
fibro.anchors <- FindIntegrationAnchors(object.list = fibro.list, dims=1:30,anchor.features = 3000,k.filter=min)
fibro.integrated <- IntegrateData(anchorset = fibro.anchors,dims=1:30)
DefaultAssay(object = fibro.integrated) <- "integrated"
fibro.integrated <- ScaleData(fibro.integrated, verbose = FALSE)
fibro.integrated <- RunPCA(fibro.integrated, npcs = 50, verbose = FALSE)
ElbowPlot(fibro.integrated,ndims=50)
fibro.integrated <- FindNeighbors(fibro.integrated, dims = 1:20)
fibro.integrated <- FindClusters(fibro.integrated, resolution = 0.1)
table(Idents(fibro.integrated))
saveRDS(fibro.integrated, file = "/group/card2/Neda/MCRI_LAB/must-do-projects/EnzoPorrelloLab/dilated-cardiomyopathy/data/fibro-int-FND.Rds")

```

```{r}
fibro.integrated <- readRDS("/group/card2/Neda/MCRI_LAB/must-do-projects/EnzoPorrelloLab/dilated-cardiomyopathy/data/fibro-int-FND.Rds")

fibro.integrated$orig.ident <- factor(fibro.integrated$orig.ident,levels = c("Fetal","ND","DCM"))
fibro.integrated$biorep <- factor(fibro.integrated$biorep,levels = c("f1","f2","f3","nd1","nd2","nd3","d1","d2","d3","d4"))
table(fibro.integrated$orig.ident)
table(fibro.integrated$biorep)

```

```{r,fig.height=10, fig.width=12}
VizDimLoadings(fibro.integrated, dims = 1:4, reduction = "pca")
```

```{r}
DimPlot(fibro.integrated, reduction = "pca",group.by="orig.ident")
DimPlot(fibro.integrated, reduction = "pca",group.by="biorep")
DimPlot(fibro.integrated, reduction = "pca",group.by="sex")
DimPlot(fibro.integrated, reduction = "pca",group.by="batch")
```

```{r,fig.height=10, fig.width=12}
DimHeatmap(fibro.integrated, dims = 1:15, cells = 500, balanced = TRUE)
DimHeatmap(fibro.integrated, dims = 16:30, cells = 500, balanced = TRUE)
DimHeatmap(fibro.integrated, dims = 31:45, cells = 500, balanced = TRUE)
```

```{r}
par(mar=c(5,4,2,2))
barplot(table(Idents(fibro.integrated)),ylab="Number of cells",xlab="Clusters")
title("Number of cells in each cluster")
```


## Visualisation with UMAP

```{r tsne}
set.seed(10)
fibro.integrated <- RunUMAP(fibro.integrated, reduction = "pca", dims = 1:20)
```


```{r}
DimPlot(fibro.integrated, reduction = "umap",label=TRUE,label.size = 6,pt.size = 0.5, split.by = "orig.ident")+NoLegend()
```


```{r}
DimPlot(fibro.integrated, reduction = "umap", group.by = "biorep")
```

```{r}
DimPlot(fibro.integrated, reduction = "umap", group.by = "sex")
```

```{r}
DimPlot(fibro.integrated, reduction = "umap", group.by = "batch")
```


```{r,fig.width=12,fig.height=4}
par(mfrow=c(1,1))
par(mar=c(4,4,2,2))
tab <- table(Idents(fibro.integrated),fibro.integrated$biorep)
barplot(t(tab/rowSums(tab)),beside=TRUE,col=ggplotColors(10),legend=TRUE)
```

```{r,fig.width=12,fig.height=4}
par(mfrow=c(1,1))
par(mar=c(4,4,2,2))
tab <- table(Idents(fibro.integrated),fibro.integrated$orig.ident)
barplot(t(tab/rowSums(tab)),beside=TRUE,col=ggplotColors(3))
legend("topleft",legend=colnames(tab),fill=ggplotColors(3))
```

# Find DEG per sub cluster

```{r}
DefaultAssay(fibro.integrated) <- "RNA"
Idents(fibro.integrated) <- fibro.integrated$integrated_snn_res.0.1
# Load unfiltered counts matrix for every sample (object all)
load("/group/card2/Neda/MCRI_LAB/single_cell_nuclei_rnaseq/Porello-heart-snRNAseq/output/RDataObjects/all-counts.Rdata")
```

```{r}
columns(org.Hs.eg.db)
ann <- AnnotationDbi:::select(org.Hs.eg.db,keys=rownames(all),columns=c("SYMBOL","ENTREZID","ENSEMBL","GENENAME","CHR"),keytype = "SYMBOL")
m <- match(rownames(all),ann$SYMBOL)
ann <- ann[m,]
table(ann$SYMBOL==rownames(all))
```

```{r}
mito <- grep("mitochondrial",ann$GENENAME)
length(mito)
ribo <- grep("ribosomal",ann$GENENAME)
length(ribo)
missingEZID <- which(is.na(ann$ENTREZID))
length(missingEZID)
```

# Find Markers

```{r markers}
# Limma-trend for DE
m <- match(colnames(fibro.integrated),colnames(all))
all.counts <- all[,m]
```

```{r}
chuck <- unique(c(mito,ribo,missingEZID))
length(chuck)
all.counts.keep <- all.counts[-chuck,]
ann.keep <- ann[-chuck,]
table(ann.keep$SYMBOL==rownames(all.counts.keep))
```

```{r}
numzero.genes <- rowSums(all.counts.keep==0)

table(numzero.genes > (ncol(all.counts.keep)-20))

keep.genes <- numzero.genes < (ncol(all.counts.keep)-20)
table(keep.genes)

all.keep <- all.counts.keep[keep.genes,]
dim(all.keep)

ann.keep <- ann.keep[keep.genes,]
```


```{r}
y.fibro <- DGEList(all.keep)

logcounts <- normCounts(y.fibro,log=TRUE,prior.count=0.5)

maxclust <- length(levels(Idents(fibro.integrated)))-1

grp <- paste("c",Idents(fibro.integrated),sep = "")
grp <- factor(grp,levels = paste("c",0:maxclust,sep=""))

design <- model.matrix(~0+grp+fibro.integrated$biorep)
colnames(design)[1:(maxclust+1)] <- levels(grp)

mycont <- matrix(0,ncol=length(levels(grp)),nrow=length(levels(grp)))
colnames(mycont)<-levels(grp)
diag(mycont)<-1
mycont[upper.tri(mycont)]<- -1/(length(levels(factor(grp)))-1)
mycont[lower.tri(mycont)]<- -1/(length(levels(factor(grp)))-1)

# Fill out remaining rows with 0s
zero.rows <- matrix(0,ncol=length(levels(grp)),nrow=(ncol(design)-length(levels(Idents(fibro.integrated)))))
test <- rbind(mycont,zero.rows)


fit <- lmFit(logcounts,design)
fit.cont <- contrasts.fit(fit,contrasts=test)
fit.cont <- eBayes(fit.cont,trend=TRUE,robust=TRUE)

fit.cont$genes <- ann.keep

summary(decideTests(fit.cont))

treat <- treat(fit.cont,lfc=0.25)

dt <- decideTests(treat)
summary(dt)

```

```{r,fig.width=10,fig.height=10}
par(mfrow=c(2,3))
for(i in 1:ncol(mycont)){
  plotMD(treat,coef=i,status = dt[,i],hl.cex=0.5)
  abline(h=0,col=colours()[c(226)])
  lines(lowess(treat$Amean,treat$coefficients[,i]),lwd=1.5,col=4)
}
```


## Write out marker genes for each cluster

```{rm eval=FALSE}
contnames <- colnames(mycont)

for(i in 1:length(contnames)){
  topsig <- topTreat(treat,coef=i,n=Inf)
  write.csv(topsig,file=paste("./output/Fib-Cluster-",contnames[i],".csv",sep=""))
}
```

```{rm}
fdr <- apply(treat$p.value, 2, function(x) p.adjust(x, method="BH"))
output <- data.frame(treat$genes,LogFC=treat$coefficients,AveExp=treat$Amean,tstat=treat$t, pvalue=treat$p.value, fdr=fdr)
write.csv(output,file="./output/Fib-MarkerAnalysis.csv")
```

# Perform gene set testing on C2 and GO sets

```{rm}
contnames <- colnames(mycont)

load("/group/card2/Neda/MCRI_LAB/single_cell_nuclei_rnaseq/Porello-heart-snRNAseq/output/RDataObjects/human_c2_v5p2.rdata")
load("/group/card2/Neda/MCRI_LAB/single_cell_nuclei_rnaseq/Porello-heart-snRNAseq/output/RDataObjects/human_c5_v5p2.rdata")

c2.id <- ids2indices(Hs.c2,treat$genes$ENTREZID)
c5.id <- ids2indices(Hs.c5,treat$genes$ENTREZID)

reactome.id <-c2.id[grep("REACTOME",names(c2.id))]

c2.c0 <- cameraPR(treat$t[,1],c2.id)
reactome.c0 <- cameraPR(treat$t[,1],reactome.id)
go.c0 <- cameraPR(treat$t[,1],c5.id)

for(i in 1:length(contnames)){
  write.csv(cameraPR(treat$t[,i],c2.id),file=paste("./output/Fib-GeneSetTests-c2-",contnames[i],".csv",sep=""))
  write.csv(cameraPR(treat$t[,i],reactome.id),file=paste("./output/Fib-GeneSetTests-reactome-",contnames[i],".csv",sep=""))
  write.csv(cameraPR(treat$t[,i],c5.id),file=paste("./output/Fib-GeneSetTests-go-",contnames[i],".csv",sep=""))
}
```

# Check quality of clusters

The quality of the clusters look good.

```{r}
par(mfrow=c(1,1))
numgenes <- colSums(all.keep!=0)
boxplot(numgenes~grp)
```


# Heatmap of marker genes

```{r}
sam <- factor(fibro.integrated$biorep,levels=c("f1","f2","f3","nd1","nd2","nd3","d1","d2","d3","d4"))
newgrp <- paste(grp,sam,sep=".")
newgrp <- factor(newgrp,levels=paste(rep(levels(grp),each=10),levels(sam),sep="."))
o <-order(newgrp)
clust <- rep(levels(grp),each=10)
samps <- rep(levels(sam),length(levels(grp)))
```

## Summarise expression across cells 

```{r}
sumexpr <- matrix(NA,nrow=nrow(logcounts),ncol=length(levels(newgrp)))
rownames(sumexpr) <- rownames(logcounts)
colnames(sumexpr) <- levels(newgrp)

for(i in 1:nrow(sumexpr)){
  sumexpr[i,] <- tapply(logcounts[i,],newgrp,mean)
}

```

```{r sigheatmap, fig.width=12,fig.height=12}
sig.genes <- gene.label <- vector("list", length(levels(grp)))
for(i in 1:length(sig.genes)){
  top <- topTreat(treat,coef=i,n=Inf)
  sig.genes[[i]] <- rownames(top)[top$logFC>0][1:10]
  gene.label[[i]] <- paste(rownames(top)[top$logFC>0][1:10],levels(grp)[i],sep="-")
} 

csig <- unlist(sig.genes)
genes <- unlist(gene.label)

myColors <- list(Clust=NA,Sample=NA)
myColors$Clust<-sample(ggplotColors(length(levels(grp))),length(levels(grp)))
names(myColors$Clust)<-levels(grp)
myColors$Sample <- sample(ggplotColors(length(levels(sam))),length(levels(sam)))
names(myColors$Sample) <- levels(sam)


#pdf(file="./output/Figures/fibro-heatmap-siggenes-summarised-FND-filtered.pdf",width=20,height=20,onefile = FALSE)
aheatmap(sumexpr[csig,],Rowv = NA,Colv = NA, labRow = genes,
         annCol=list(Clust=clust,Sample=samps),
         annColors=myColors,
         fontsize=16,color="-RdYlBu",
         scale="none")
#dev.off()
```










